---
title: "Analyzing single-cell RNA-seq perturbation with DECAL"
author: "André M. Ribeiro-dos-Santos"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  TODO: Package abstract
  decal package version: `r packageVersion("decal")`
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    df_print: kable
    fig_width: 5
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Analyzing single-cell RNA-seq perturbation with scCloneDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  tidy = FALSE, cache = FALSE, dev = "png", collapse = TRUE, comment = "#>",
  message = FALSE, error = FALSE, warning = TRUE
)
options(scipen = 1, digits = 2)
```

# DECAL Workflow

**DECAL** (**D**ifferential **E**xpression analysis of **C**lonal
**A**lterations **L**ocal effects) provide you tools to conduct differential
expression analysis of single-cell perturbations to potential interacting
genes.
Similar to other expression analysis tools, it models gene expression using
a _Negative Binomial_ (or _Gamma-Poisson_) regression, modeling each gene
UMI count by the cell total count and the cell alteration status.

## Quick Start

`decal` is the package main function which estimates each gene dispersion
($\theta$) parameter and fit a _Negative Binomial_ regression to each gene
and alteration pair specified to evaluate the perturbation statistical
significance.
It requires three parameters a UMI count matrix, a list specifying your clones
cell (count matrix columns) composition, and a table indicating the potential
alteration and affected gene to evaluate.

This package provide you tools to conduct differential expression analysis of
single-cell perturbations to interacting genes.
It models the perturbation to a gene UMI count using a _Negative Binomial_
regression controlled by the cell total count.
Provided the required data, the function `decal` will estimate each gene
dispersion parameter and fit a regression model to each gene and perturbation
pair and evaluate the perturbation statistical significance.
Such analysis is usually conducted after several steps which include generating
your single-cell UMI count matrix, determination of your perturbations clonal
population, and electing potential perturbations to evaluate.

Below we show a quick example of the basic steps to conduct the interactions
differential expression analysis using our package.

```{r quick_start}
library(decal)

data("sim_decal")
perturbations <- sim_decal$perturbations
count <- sim_decal$count
clone <- sim_decal$clone

res <- decal(perturbations, count, clone)
head(res)
```

Here we used a simulated dataset (`sc_simulated`) made available and is
composed of the three required parameters: a UMI count matrix (`count`), a
list of cells dived by clonal population (`clone`), and a table of candidate
interactions to be evaluated (`perturbations`).
Given the required input, `decal` conducts the whole differential analysis
and updates the `perturbations` table to include the following:

- `n0` and `n1`: number of non-perturbed and perturbed cells, respectively.
- `x0` and `x1`: average UMI count among perturbed and non-perturbed cells,
  respectively.
- `mu`: average UMI count among all cells.
- `xb`: expected average UMI count of perturbed cells.
- `theta`: estimated _negative binomial_ dispersion parameter.
- `z`: estimated perturbation z-score.
- `lfc`: _log2 fold-change_ of perturbed cells gene expression.
- `pvalue` and `p_adjusted`: perturbation _t-test_ significance values.

## Features

## XX

Admitting a _false discovery rate_ (_FDR_) of 05%, we can explore which
interactions indicated real expression change by picking those with
`p_adjusted < 0.05`. See below the results for top 5 significant interactions
with increasing and decreasing expression effects.

```{r quick_start_significant}
sig <- subset(res, p_adjusted < 0.05)
sig <- sig[order(sig$lfc), ]

sig[c(1:5, nrow(sig) - 1:5),]
```

```{r quick_start_plot_it, fig.height = 4}
sf <- mean(colSums(count)) / colSums(count)

par(mfrow = c(2, 2), mar = c(2, 2, 3, 2))
for (i in c(1, 2, nrow(sig) - 1, nrow(sig) - 2)) {
  x <- colnames(count) %in% clone[[sig$clone[i]]]
  y <- log2(count[sig$gene[i],] * sf + 1)
  n <- length(x)
  title <- paste0(sig$clone[i], "+", sig$gene[i],"\nLFC: ", round(sig$lfc[i], 2))
  plot(
    x + runif(n, -.3, .3), y, xlab = "", ylab = "Gene Expression",
    axes = FALSE, xlim = c(-.5, 1.5), ylim = c(0.9, 1.1) * range(y),
    main = title
  )
  axis(1, at = 0:1, labels = c("perturbed", "unperturbed"))
  axis(2, at = log2(c(0, 5, 10, 20, 50, 100) + 1), labels = c(0, 5, 10, 20, 50, 100))
  ## add error bar
  ymed <- c(sig$mu[i], sig$xb[i])
  ylow <- log2(qnbinom(0.025, size = sig$theta[i], mu = ymed) + 1)
  yupr <- log2(qnbinom(0.975, size = sig$theta[i], mu = ymed) + 1)
  points(c(0, 1), log2(ymed + 1), col = "firebrick", cex = 1, pch = 16)
  arrows(
    x0 = c(0, 1), y0 = ylow, x1 = c(0, 1), y1 = yupr,
    code = 3, angle = 90, col = "firebrick", length = .1, lwd = 2
  )
}
```

Since we include some real gene perturbation to our simulated dataset, we can
explore how well our model predicts the change. The `perturbations` table
includes a column `expected_lfc` column that indicates the real perturbation
applied when simulating `count`, where `expected_lfc == 0` indicate no
perturbation was applied.
First, we can evaluate our model fitting using the following confusion matrix.

```{r quick_start_confusion}
real_effect <- res$expected_lfc != 0
estimated_effect <- res$p_adjusted < 0.05

confusion_mat <- table(estimated_effect, real_effect)
round(prop.table(confusion_mat), 3)
```
The confusion matrix indicates that admitting 5% _FDR_, our model presented:

- **Accuracy**: `r round(sum(diag(confusion_mat)) / sum(confusion_mat), 3)`
- **Precision**: `r round(confusion_mat[2, 2] / sum(confusion_mat[2,]), 3)`
- **Recall**: `r round(confusion_mat[2, 2] / sum(confusion_mat[,2]), 3)`
- **Observed FDR**: `r round((confusion_mat[1, 2] + confusion_mat[2, 1]) / sum(confusion_mat), 3)`

We can also evaluated how well our model predicted the real _log2 fold-change_
perturbation.
Illustrated below is the estimated `lfc` distribution for each value of
perturbation applied.

```{r quick_start_hist}
real_lfc <- res$expected_lfc
estimate_lfc <- res$lfc

breaks <- c(-Inf, seq(-5, 5, length.out = 30), Inf)
histograms <- lapply(split(estimate_lfc, real_lfc), hist, breaks = breaks, plot = FALSE)

ymax <- max(sapply(histograms, function(x) x$count / sum(x$count)))
plot(
  x = c(-5, 5), y = c(0, ymax), type = "n",
  xlab = "Estimated LFC", ylab = "Proportion of interactions"
)
for (i in seq_along(histograms)) {
  hist <- histograms[[i]]
  lines(hist$mids, hist$count/sum(hist$count), col = i, type = "s")
}
legend(
  "topleft", title = "Real LFC", legend = names(histograms),
  col = seq_along(histograms), bty = "n", lty = 1, lwd = 2, ncol = 2
)
```

Below we measured the estimate _Mean Squared Error_ (_MSE_) and _Rooted MSE_
(_RMSE_) for each real perturbation value.

```{r quick_start_rmse}
mse_estimate <- sapply(split(res, res$expected_lfc), function(dat) {
  dat <- dat[complete.cases(x),]
  mse <- mean( (dat$expected_lfc - dat$lfc)^2, na.rm = TRUE )
  c(real_lfc = dat$expected_lfc[1], N = nrow(x), MSE = mse, RMSE = sqrt(mse))
})

round(t(mse_estimate), 2)
```

Though the _MSE_ increased for negative _LFC_, the model presents high accuracy
and precision in estimating real perturbation effects as demonstrated by the
results above.

## Understanding arguments

The analysis is performed by the function `decal` which requires three main
parameters:

1. A table of clones and potentially perturbed genes (or feature).
2. A single-cell UMI count matrix.
3. A list of cells defining each clone composition.

Here we will dive down into 1 and 3, and how to define them.

### Defining your cells clonal structure

The analysis was designed to explore the expression difference of a
perturbation replicated to a clonal population of cells. We recommend using
an adjacent dataset to define your cells clonal populations and their
perturbation composition. For the analysis here, we recommend define each
of your clones' cell composition as a list of character (cells id) or integer
(cells index) vectors, such as the example below:

```{r var_clone}
clone[1:3]
```

### Electing potential gene perturbation

Since we aim to explore specific gene perturbations, we must specify pairs of
clone population and genes (or features) to evaluate expression change among
the clone cells in regards to all others.
In our original study, we located our perturbations and evaluated expression
change to all genes with TSS within _250Kbp_ of the perturbation.
Here, we specify the these pairs of clone and perturbed genes to evaluate as
a table with a `clone` and `gene` column indicating the clone/gene id
(as character) or index (as integer), such as the example below:

```{r var_interactions}
head(perturbations)
```

## Reducing noise and increasing power

To improve results, `decal` include some filtering step to skip testing
interactions with low statistical power.
Among these filter parameters are included:

- `min_x`: minimal average count observed on non-perturbed or perturbed cells
  (indicated by `x0` and `x1`, respectively, on the output table).
- `min_n`: minimal number of perturbed cells (indicated by `n1`).
- `min_mu`: minimal global average count (indicated by `mu`), also
  required to estimate dispersion`theta`.

Interactions that doesn't met all of theses requirements are skipped and no
differential analysis is conducted, though they are still included in the
result table.
By default, `decal` is set to our suggestion that presented a good
performance in our dataset: `min_x = 1`; `min_n = 2`; and `min_mu = 0.05`.

# Understanding decal model

## Statistical model

The `decal` model is based on the observations of _Svensson (2020)_ and
_Townes et al. 2019_ that the UMI counts of a particular gene approximates a
_Poisson_ or _Negative Binomial_ distribution in a single-cell RNA-seq
experiment. Thus, we proposed the observed count with following model:

$$Y_{gc} \sim NB(\mu_{gci}, \theta_g)$$
$$log(\mu_{gci}) = \beta_g + \beta_d D_c + \beta_x X_{ci}$$

where UMI count $Y_{cg}$ for gene _g_ and cell _c_ are modeled using a
_negative binomial_ distribution with fitted mean $\mu_{cgi}$ and a
gene-specific dispersion parameter $\theta_g$. The log fitted mean is
proportional to the cell log total count ($D_c = \sum_g Y_{cg}$) and the
perturbation effect ($\beta_x$) of a specific clone _i_ with $X_{ci}$ is
an indicator variable if the cell belong or not the perturbed clone.

The parameter $\theta_g$ determines the model variance where smaller values
indicate a wider distribution and higher values produce a tighter distribution
that coincides with a Poisson distribution.

$$Var(Y_{cg}) = E[(Y_{cg} - \mu_{cgi})^2] = \mu_{cgi} + \mu_{cgi}^2/\theta_g$$

$$\lim_{\theta\to\infty} NB(\mu, \theta) = Pois(\mu)$$

Our model is similar to previously published tools such as`edgeR`
(_Robinson et al. 2010_; _McCarthy et al. 2012_), `DESeq`
(_Anders and Huber, 2010_; _Love et al. 2014_), and glmGamPoi
(_Ahlmann-Eltze & Huber 2020_).
But it differs in two points: (i) we adopted a regularized strategy to estimate
$\theta_g$ to make it more robust to sampling noise such as described by
_Hafemeister & Satija (2019)_; and (ii)

- TODO: complete section

## Estimating dispersion

Here we apply the strategy used by _Hafemeister & Satija (2019)_ to estimate
$\theta_g$. First, we sample ~ 2000 genes and make a naive estimate of their
expression using a Poisson regression modeled as $Y_{cg} \sim Poi(mu_{cg})$
such as $log(mu_{cg}) = \beta_g + \beta_d D_c$. Next, we estimate a naive
dispersion parameter using a maximum likelihood estimator, fit a kernel smooth
regression to regularize the parameter as a function of average gene count
($\sum_g Y_{cg} / N$), and using this regression to produce the final $\theta_g$
for all other genes.

```{r}
dat <- res[, c("gene", "mu", "theta", "raw_theta")]
dat <- res[complete.cases(res),]
dat <- unique(dat[order(dat$mu),])

plot(
  log10(dat$mu), dat$raw_theta,
  xlab = "Average expression (mu)", ylab = expression(theta),
  xaxt = "n", ylim = c(0, 150)
)
abline(h = 100)
lines(log10(dat$mu), dat$theta, col = "firebrick")
legend(
  "topleft", c("naive estimate", "regularized estimate", "real"),
  pch = c(1, NA, NA), lty = c(NA, 1, 1), col = c(1, 2, 1)
)
breaks <- log10(c(0.01, 0.1, 1, 2, 5, 10, 20, 50, 100))
axis(1, at = breaks, labels = round(10**breaks, 2))
```

# Simulating
xx

Many factor can affect our model **power** and **sensitivity**, thus we also
include tools to simulate and measure your experiments power under your
conditions

Thus we can measure our false discovery rate, given our current cutoff
(`qvalue < 0.10`), by producing the following confusion matrix.

```{r quick_start_confusion_matrix}
# round(prop.table(
#   table(real = res$estimated_lfc != 0, estimated = res$p_adjusted < 0.05)), 2)
```

To power your analysis, I've included some functions that facilitate and allow
you to perform a randomization experiment to estimate the test `z-score` under
no effect and estimate test power under our model assumptions.

## Null hypothesis estimation

```{r sim_randomization, fig.height=3, fig.width=9}
rnd_perturbations <- data.frame(
  clone = sample(names(clone), 1000, replace = TRUE),
  gene = sample(rownames(count), 1000, replace = TRUE)
)
rnd_res <- decal(rnd_perturbations, count, clone)
rnd_res <- subset(rnd_res, !is.na(rnd_res$pvalue))

par(mfrow = c(1, 4))
hist(rnd_res$z, xlab = "Z-score", main = "")
hist(rnd_res$pvalue, xlab = "Observed Pvalue", main = "")
plot(-log10(ppoints(rnd_res$pvalue)), -log10(sort(rnd_res$pvalue)),
  xlab = "Expected -log10(P)",
  ylab = "Observed -log10(P)")
abline(0, 1, col = "red")
hist(rnd_res$p_adjusted, xlab = "Observed Qvalue", main = "")
```

## Power analysis

```{r sim_power, fig.height=4, fig.width=8}
## for each clone perturb 10 genes with -2, -1, 1, and 2 log2 FC and add 100
## interactions with no effect.
lfc <- c(-2, -1, 1, 2)
pwr_lfc <- c(rep(lfc, each = 10), rep(0, 100))
pwr_dat <- sim_experiment_from_data(count, pwr_lfc, nclones = 20, min_n = 5, max_n = 20)
dimnames(pwr_dat$count) <- NULL
pwr_res <- decal(pwr_dat$perturbations, pwr_dat$count, pwr_dat$clone)
pwr_res <- subset(pwr_res, !is.na(pvalue))

## Confusion matrix
prop.table(table(
  actual = pwr_res$expected_lfc != 0,
  estimated = pwr_res$p_adjusted < 0.10
  ),2)

## Histogram of estimated LFC for different effects
breaks <- c(-Inf, seq(-3.5, 3.5, length.out = 30) ,Inf)
histograms <- lapply(
  split(
    pwr_res$lfc[pwr_res$expected_lfc %in% lfc],
    pwr_res$expected_lfc[pwr_res$expected_lfc %in% lfc]
  ), hist, breaks = breaks, plot = FALSE
)
ymax <- max(sapply(histograms, function(x) x$count / sum(x$count))) * 1.05
## Estimate power by LFC
pwr <- sapply(split(
  pwr_res$p_adjusted[pwr_res$expected_lfc %in% lfc] < 0.05,
  pwr_res$expected_lfc[pwr_res$expected_lfc %in% lfc]
  ), mean)

par(mfrow = c(1, 2))
plot(c(-3.5, 3.5), c(0, ymax), type = "n", xlab = "Estimated LFC", ylab = "Frequency")
for(i in seq_along(histograms)) {
  lines(histograms[[i]]$mids, histograms[[i]]$count / sum(histograms[[i]]$count), col = i, type = "s")
}
legend(
  "topleft", title = "Actual LFC", legend = sprintf("%+d", lfc),
  col = seq_along(lfc), bty = "n", lty = 1, cex = .7
)
## Barplot of power estimate
barplot(pwr, xlab = "LFC", ylab = "Power")
```

# Assumptions and Caveats

xxx

# Features

- **tidyverse** compatible

# Session info

```{r sessionInfo}
sessionInfo()
```

# References

