---
title: "Analyzing single-cell RNA-seq perturbation with DECAL"
author: "André M. Ribeiro-dos-Santos"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: >
  Single-cell RNA sequencing technology has allow us to explore cell specific
  expression patterns. Functional genomics have been exploring such dataset to
  establish direct link between genetic alterations and differential gene
  expression. The package DECAL (Differential Expression analysis of Clonal
  Alterations Local effects) provides tools to explore single-cell of
  perturbed cells populations. It uses a negative binomial generalized linear
  model to evaluate gene expression changes between clonal cell populations and
  gene candidates. The present vignette explains the package usage,
  demonstrantes a typical analysis workflow, and explains the package
  statistical model.
  decal package version: `r packageVersion("decal")`
output:
  bookdown::html_document2:
    highlight: pygments
    toc: true
    df_print: kable
    fig_width: 6
    fig_caption: true
    number_sections: false
  bookdown::pdf_document2: default
pkgdown:
  as_is: true
bibliography: decal.bib
vignette: >
  %\VignetteIndexEntry{Analyzing single-cell RNA-seq perturbation with DECAL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  tidy = FALSE, cache = FALSE, dev = "png", collapse = TRUE, comment = "#>",
  message = FALSE, error = FALSE, warning = TRUE
)
options(scipen = 1, digits = 2)
```

# Default Workflow

**DECAL** (**D**ifferential **E**xpression analysis of **C**lonal
**A**lterations **L**ocal effects) provide you with tools to conduct
differential expression analysis of single-cell perturbations experiments
to potentially interacting genes.
Similar to other differential expression analysis tools, it models gene
expression using a _Negative Binomial_ (or _Gamma-Poisson_) regression,
modeling each gene UMI count by the cell alteration status and cell total
count.

## Features

- `decal` is compatible with **tidyverse** analysis.
- Includes a suit of simulation functions to generate your own dataset based
  `decal` model and evaluate your statistical power.
- Package has few dependencies, requiring only `MASS`, `fastglm` and `Matrix`.
- Can evaluate specific clonal alteration and gene effect, instead of
  modeling all genes and all alterations. This allow us to quickly investigate
  a large number of interactions skipping unlikely effects.

## Installation

To install `decal` package current version, open your R terminal and type:

```{r installation, eval = FALSE}
## Install remotes if not available with
## install.install.packages("remotes")
remotes::install_github("mauranolab/decal")
```

## Using `decal`

The package main function (`decal`) runs the whole statistical analysis by
fitting a _Negative Binomial_ (or _Gamma-Poisson_) regression to each gene
and alteration pair specified and evaluate the perturbation statistical
significance for a single-cell perturbation experiment. You can run `decal`
as follow:

```{r example, eval = FALSE}
decal(perturbations, count, clone)
```

The three parameters required are the following:

- a table specifying the population and gene to evaluate perturbation
  (`perturbations`).
- a UMI count matrix that each column correspond to a cell and each row
  a gene or feature (`count`).
- a list of cells specifying your clones composition (`clone`). This is
  represented by a R list composed by character (or integer) vectors named
  after the clone ID and indicating the cells belonging to this clone.

`decal` returns a deep copy of the `perturbations` table with the following
additional columns:

- `n0` and `n1`: number of non-perturbed and perturbed cells, respectively.
- `x0` and `x1`: average UMI count among perturbed and non-perturbed cells,
  respectively.
- `mu`: average UMI count among all cells.
- `xb`: expected average UMI count of perturbed cells.
- `theta`: estimated _negative binomial_ dispersion parameter.
- `z`: estimated perturbation z-score.
- `lfc`: _log2 fold-change_ of perturbed cells gene expression.
- `pvalue` and `p_adjusted`: perturbation _t-test_ significance values.

## Quick Start

Below we include a example of `decal` analysis on a simulated dataset included
with our package.

```{r quick_start}
library(decal)

data("sim_decal")
perturbations <- sim_decal$perturbations
count <- sim_decal$count
clone <- sim_decal$clone

res <- decal(perturbations, count, clone)
head(res)
```

## Exploring decal results

Now let's explore `decal` results. Admitting a 5% _false discovery rate_
(_FDR_), we can select significantly differentiate perturbations by
picking those with `p_adjusted < 0.05`.
Below we present the top 5 interactions with highest reducing and increasing
log fold-change effect (as measured by `lfc`).

```{r quick_start_significant}
sig <- subset(res, p_adjusted < 0.05)
sig <- sig[order(sig$lfc), ]
## select first 5 and last 5 since`sig` is ordered by lfc
pick <- c(1:5, nrow(sig) - 1:5)
sig[pick,]
```

To visualize the perturbation effect, let's plot the top 3 interactions
increasing and reducing gene expression.

```{r quick_start_plot, fig.height = 4}
## compute a normalize size-factor
sf <- mean(colSums(count)) / colSums(count)
## add log10(x+1) transformation function
log10p <- function(x) log10(x + 1)

## pick tables first 3 and last 3
pick <- c(1:3, nrow(sig) - 1:3)

par(mfrow = c(2, 3))
for (i in pick) {
  ## Determinate perturbed and unperturbed cells
  x <- colnames(count) %in% clone[[sig$clone[i]]]
  ## Calculate normalized expression count
  y <- count[sig$gene[i],] * sf
  ## plot jitter
  title <- paste0(sig$clone[i], "+", sig$gene[i],"\nLFC: ", round(sig$lfc[i], 2))
  plot(
    x + runif(length(x), -.3, .3), log10p(y), main = title,
    axes = FALSE, xlab = "", ylab = "Normalize UMI count",
    xlim = c(-.5, 1.5), ylim = c(0.9, 1.1) * range(log10p(y))
  )
  axis(1, at = 0:1, labels = c("perturbed", "unperturbed"), las = 2)
  axis(2, at = log10p(c(0, 2, 5, 10, 20, 50, 100)), labels = c(0, 2, 5, 10, 20, 50, 100))
  ## add 95% error bar
  ymed <- c(sig$mu[i], sig$xb[i])
  ylow <- qnbinom(0.025, size = sig$theta[i], mu = ymed)
  yupr <- qnbinom(0.975, size = sig$theta[i], mu = ymed)
  points(c(0, 1), log10p(ymed), col = "firebrick", cex = 2, pch = 16)
  arrows(
    x0 = c(0, 1), y0 = log10p(ylow), x1 = c(0, 1), y1 = log10p(yupr),
    code = 3, angle = 90, col = "firebrick", length = .1, lwd = 2
  )
}
```

## Evaluating decal results

In our `sim_decal` simulated dataset, we introduced some real expression
perturbations as indicated by `expected_lfc` column in `perturbations` table,
such as a 0 `expected_lfc` indicate no change was applied.
We can evaluate our results using the confusion matrix below:

```{r quick_start_confusion}
conf_mat <- table(real = res$expected_lfc != 0, decal = res$p_adjusted < 0.05)
round(prop.table(conf_mat), 3)
```

Based on this confusion matrix, our `decal` model presented:

- **Accuracy**: `r round(sum(diag(conf_mat)) / sum(conf_mat), 3)`
- **Precision**: `r round(prc <- conf_mat[2, 2] / sum(conf_mat[2,]), 3)`
- **Recall**: `r round(rec <- conf_mat[2, 2] / sum(conf_mat[,2]), 3)`
- **F1 Score**: `r round(2 * prc * rec / (prc + rec) , 3)`
- **Observed FDR**: `r round((conf_mat[1, 2] + conf_mat[2, 1]) / sum(conf_mat), 3)`

These results show great metrics with recall, accuracy, precision and F1 score
above 95% and a _false discovery ratio_ within our admitted 5%.

Next, let's explore how well `decal` estimated the real _log2 fold-change_ (LFC)
applied. Illustrated below is the decal estimated LFC (`lfc` column in `res`)
distribution for each of the real perturbations values applied (`expected_lfc`
column).

```{r quick_start_hist}
lfc_histogram <- function(lfc, fn = function(x) { x$count / sum(x$count) }) {
  hist_breaks <- c(-Inf, seq(-4, 4, length.out = 50), Inf)
  hist_counts <- lapply(lfc, hist, breaks = hist_breaks, plot = FALSE)
  hist_ymax <- max(sapply(hist_counts, fn))
  plot(c(-4, 4), c(0, hist_ymax), type = "n",
    xlab = "Estimated LFC", ylab = "Proportion of interactions")
  for(i in seq_along(hist_counts)) {
    lines(hist_counts[[i]]$mids, fn(hist_counts[[i]]), col = i, type = "s")
  }
  legend("topleft", title = "Real LFC", legend = names(hist_counts),
    col = seq_along(hist_counts), bty = "n", lty = 1, lwd = 2, ncol = 2)
}

lfc_histogram(split(res$lfc, res$expected_lfc))
```

Finally, we can measure decal LFC estimative _Mean Squared Error_ (_MSE_) and
_Rooted MSE_ (_RMSE_) using the code below. The results indicate that
`decal` estimates presented high accuracy and precision with _RMSE_ below 1 LFC
for all perturbations effects applied.

```{r quick_start_rmse}
err <- (res$lfc - res$expected_lfc)**2
err <- split(err, res$expected_lfc)
mse <- sapply(err, mean, na.rm = TRUE)

data.frame(
  N = sapply(err, length),
  MSE = mse, RMSE = sqrt(mse)
)
```

# Exploring `decal` arguments

As mentioned before, `decal` function requires three main parameters:

1. `perturbation`, a table specifying the clone population and gene to
  evaluate differential expression.
2. `count`, a UMI count matrix where each column correspond to a cell and
  each row a gene or feature.
3. `clone`, a list of cells specifying each clone populations composition.

In the next sections, we will further illustrate and give alternative formats
for (1) and (3).

## Electing potential gene perturbations

Since `decal` was designed to explore specific gene perturbations, it requires
you to specify clone and gene (or features) pairs to evaluate expression change
among the cells within the clone in regards to all other cells.
For example, in our original study after we localized our alternations and
evaluated their impact to all genes within _250 kbp_.

In `sim_decal$perturbations` data.frame (see snippet below) the clone and gene
pairs are specified by the `clone` and `gene` columns, which can be an id
(as character vector) or index (as integer vector). In the case these elements
are specified as an index, they identify the ith element of `sim_decal$clone`
and row position in `sim_decal$count`, respectively.

```{r var_interactions}
head(sim_decal$perturbations)
```

By default `decal` uses the `clone` and `gene` columns, but you can specify
a different column using `gene_col` and `clone_col` parameters like in the
example below.

```{r}
alt_pert <- data.frame(gene_ix = 101:110, clone_ix = 1:10)
decal(alt_pert, count, clone,
  gene_col = "gene_ix", clone_col = "clone_ix")
```

## Defining your experiment clonal structure

After identifying your cells' clonal populations you can specify it `decal`
(`clone` parameter) as a R list where each element is clone represented by a
collection of ids (as character vector) or index (as integer vector)
indicating the set of cells belonging to the clone. When the cells are
specified by index, they correspond to the column position in your `count`
matrix. See below a example from `sim_decal$clone`

```{r var_clone}
sim_decal$clone[1:2]
```

## Reducing noise and increasing power

To improve your findings, `decal` includes some filtering arguments to avoid
testing interactions with low statistical power:

- `min_x`: minimal average UMI count on perturbed and unperturbed cells
  (indicated by `x1` and `x0` on resulting table, respectively).
- `min_n`: minimal number of perturbed cells (indicated by `n1`).
- `min_mu`: minimal global average count (indicated by `mu`), also skips
  `theta` estimation.

Interactions that doesn't met all of these requirements are skipped and no
differential expression analysis is conducted, eventhough they are no removed
from the results.
By default, `decal` applies the following minimal filters `min_x = 1`;
`min_n = 2`; and `min_mu = 0.05`.

# Understanding `decal` model

## Statistical model

The `decal` statistical model is based on the observations that UMI counts
of a particular gene approximates a _Poisson_ or _Negative Binomial_
distribution in a single-cell RNA-seq experiment [@svensson_droplet_2020;
@townes_feature_2019].
Thus, we proposed to model the observed counts as follow:

$$
Y_{gc} \sim NB(\mu_{gci}, \theta_g)
(\#eq:model)
$$

where $Y_{cg}$ is the UMI count for gene _g_ and cell _c_ modeled using a
_negative binomial_ distribution with expected count $\mu_{cgi}$ and a
gene-specific dispersion parameter $\theta_g$.

$$
log(\mu_{gci}) = log(D_c) + \beta_g + \beta_x X_{ci}
(\#eq:modelMu)
$$

The $log(\mu_{cgi})$ (or log expected count) is proportional to the
perturbation effect ($\beta_x$) of a specific clone _i_ indicated by $X_{ci}$
(where $X_{ci} == 1$ if cell _c_ belongs to the perturbed clone or
$X_{ci} == 0$ otherwise) and offseted by the log cell total UMI count
($D_c = \sum_g Y_{cg}$).

The parameter $\theta_g$ defines the model dispersion where smaller values
produce a wider distribution and higher values produce a tighter distribution
that coincides with a _Poisson_ distribution. Such as the model variance is
defined as follow:

$$
\begin{aligned}
Var(Y_{cg}) & = E[(Y_{cg} - \mu_{cgi})^2] \\
 & = \mu_{cgi} + \mu_{cgi}^2/\theta_g \\
\lim_{\theta\to\infty} NB(\mu, \theta) & = Pois(\mu)
\end{aligned}
(\#eq:variance)
$$

This modeling approach is similar to previously published models such as edgeR
[@robinson_edger_2010; @mccarthy_differential_2012], DESeq2
[@anders_differential_2010; @love_moderated_2014], and glmGamPoi
[@ahlmann-eltze_glmgampoi_2021].
But it differs in two points: (i) to estimate $\theta_g$ we adopted a
regularized estimation strategy to make it robust to sampling noise in low
expression genes; and (ii) instead of fitting the model to all genes, it fits
and measure specific perturbation to gene.

## Estimating dispersion

To estimate and regularize our $\theta_g$, we adopted the strategy described
by @hafemeister_normalization_2019.
This strategy consists in first selecting about 2000 genes, estimating their
expect count ($mu_{cg}$) using a naive Poisson regression offseted by the log
cell total count, such as:

$$
\begin{aligned}
Y_{cg} & \sim Pois(mu_{cg}) \\
log(mu_{cg}) & = \beta_g + log(D_c)
\end{aligned}
(\#eq:naiveTheta)
$$

Next, we use $mu_{cg}$ to make a naive estimate of $\theta_g$ using a maximum
likelihood estimator (`MASS::theta_ml` function).
The resulting naive estimate was finally regularized by fitting it to a kernel
smooth regression as a function of the average gene count ($\sum_g Y_{cg} / N$)
to produce our final estimate of $\theta_g$.
Illustrate below is a scatterplot of each naive estimate, in red our final
regularized estimate and in black the real dispersion applied when simulating
`sim_decal` dataset.

```{r}
raw_theta <- attr(res, "raw_theta")
mu <- rowMeans(count)
mu_breaks <- c(0.01, 0.1, 1, 2, 5, 10, 20, 50, 100)

plot(log10(mu), raw_theta,
  xlab = "Average expression (mu)", xaxt = "n",
  ylab = expression(theta), ylim = c(0, 150))
abline(h = 100)
lines(log10(res$mu[order(res$mu)]), res$theta[order(res$mu)], col = 2, lwd = 3)
axis(1, at = log10(mu_breaks), labels = mu_breaks)
legend(
  "topleft", c("Naive Estimate", "Regularized Estimate", "Real"),
  pch = c(1, NA, NA), lty = c(NA, 1, 1), col = c(1, 2, 1)
)
```

As shown above, our regularized estimates approximates the real dispersion
values and control some of the excessive fluctuation presented by our naive
estimate.

## Assumptions and Caveats

It is worth notice that `decal` statiscal model implies some assumptions and
caveats:

1. The UMI count distribution is independent between genes, which is usually
  true for single-cell experiments due to large quantity of genes and small
  proportion of counts each of them represent.
2. The effect of individual clones are not enough to affect the unperturbed
  estimates for other clones.
  For such assumption to hold, the number of perturbated cells of any clone
  must be much smaller than the total number of cells available (`n1 << N`).
  This allow us to explore each clone independently, since it makes the effect
  of other clones to our unperturbed estimates negletable.
3. The cells investigates must represent an uniform population differing only
  by the induced alterations.
  This can safely be assumed when the experiment is conducted in a uniform
  cell culture.

# Simulating

Since many factor can affect your experiment, we included to `decal` a way to
simulate a dataset under our model assumptions (see equation \@ref(eq:model)).
These tools will allow you to estimate your experiment **power** and
**sensitivity** under different conditions you may find.

`decal` simulation suite is compose of various functions to generate a count
matrix (`sim_count` and `sim_count_from_data`), randomly assign cells to
different clones (`sim_clone` and `sim_clone_range`), and generate a full
experiment dataset (`sim_experiment` and `sim_experiment_from_data`).
Here we will use `sim_experiment` and `sim_experiment_from_data` to generate
our examples, but the other functions behave similarly.

## Evaluating our model under null hypothesis

Under the null hypothesis, we expect that our perturbations produce no
expression change to nearby genes.
Admitting a _FDR_ of 5% and that `decal` statistical model holds, we would
expect that the resulting z-score present a normal distribution
centered at 0, p-value have a uniform distribution and about 5% of the tests
below our significance threshold.

Given an existing dataset, we can evaluate if our model holds by randomly
recombining clone and gene pairs.

```{r sim_randomization, fig.height=2.5, fig.width=8}
rnd <- unique(data.frame(
  clone = sample(perturbations$clone),
  gene = sample(perturbations$gene)
))
## remove previous clone + gene pairs
rnd <- rnd[!paste(rnd$gene, rnd$clone) %in% paste(perturbations$gene, perturbations$clone),]
rnd_res <- decal(rnd, count, clone)
rnd_res <- subset(rnd_res, !is.na(rnd_res$pvalue))

par(mfrow = c(1, 3))
hist(rnd_res$z, xlab = "Z-score", main = "")
hist(rnd_res$pvalue, xlab = "Observed Pvalue", main = "")
hist(rnd_res$p_adjusted, xlab = "Observed Qvalue", main = "")

mean(rnd_res$pvalue < 0.05)
mean(rnd_res$p_adjusted < 0.05)
```

As expected, this experiment z-score presented the expected _z-score_ and
_p-value_ distribution and the ratio of tests with p-value below the
significance threshold (`r mean(rnd_res$pvalue < 0.05)`) within the expected
5% at random.

We can conduct a similar experiment producing a random count matrix based
on your experiment counts. `sim_count_from_data` will simulate a UMI count
matrix with the same dimensions as your experiment with cells total UMI count
and gene average expression similar to your reference count matrix.

```{r, fig.height=2.5, fig.width=8}
rnd_count <- sim_count_from_data(count)

rnd_res <- decal(perturbations, rnd_count, clone)
rnd_res <- subset(rnd_res, !is.na(rnd_res$pvalue))

par(mfrow = c(1, 3))
hist(rnd_res$z, xlab = "Z-score", main = "")
hist(rnd_res$pvalue, xlab = "Observed Pvalue", main = "")
hist(rnd_res$p_adjusted, xlab = "Observed Qvalue", main = "")
```

## Estimating your experiment power

Another usufull application of `decal` simulation suite is conducting a power
analysis to decide how much should you sequence or to define your tests filters.
In the example below, we generated an experiment using
`sim_experiment_from_data` which generates a random count matrix, a clone
assignment list, and a perturbation table.
The generated count matrix is based on your experiment matrix with the same
dimensions, similar cells total count and gene average expression.
The clone list is composed of 100 clones with 5 to 20 cells each.
Perturbation table indicates pairs of clone and genes and the log2 fold-change
applied, where for each clone a gene was randomly to apply a effect indicated
by `lfc`. In this case, for each clone, 180 genes were selected and listed in
`pwr_dat$perturbations` and among them 20 were perturbed by a log2 fold-change
of -2, -1, 1, and 2.

```{r power, fig.height=3, fig.width=8}
pwr_lfc <- c(rep(c(-2, -1, 1, 2), each = 20), rep(0, 100))
pwr_dat <- sim_experiment_from_data(
  count, lfc = pwr_lfc, nclones = 100, min_n = 5, max_n = 20
)
pwr_res <- decal(pwr_dat$perturbations, pwr_dat$count, pwr_dat$clone)
pwr_res <- subset(pwr_res, !is.na(pvalue))
```

Adminitting a 5% FDR, we can estimated the ratio of the truly perturbated
gene and clone pairs were detected by `decal` (left plot) and the clone
size frequency (right plot).
For all log2 fold-change applied here, `decal` obtained a power close to 80%
that indicate que performance of the model.

```{r power_plot, fig.height=3, fig.width=8}
par(mfrow = c(1, 2))
barplot(
  sapply(split(
    pwr_res$p_adjusted[pwr_res$expected_lfc != 0] < 0.05,
    pwr_res$expected_lfc[pwr_res$expected_lfc != 0]
  ), mean),
  xlab = "LFC", ylab = "Power")
abline(h = 0.8, lty = 2, col = "firebrick")
hist(sapply(pwr_dat$clone, length), main = "", xlab = "Number of perturbed cells")
```

# References

<div id="refs"></div>

# Session info

```{r sessionInfo}
sessionInfo()
```

```{css, echo=FALSE}
p.abstract{
  font-size: 24px;
  font-family: inherit;
  text-align: center;
  font-weight: bold;
}

div.abstract{
  margin: auto;
  width: 90%;
  padding-bottom: 15px;
}
```
